Nesterov's First Order Method Accelerations: Unifications, Applications and Numerical Experiments

### 2025-03-04 Notebook LaTEX/Proposal/presentation.tex
>w<, changes

% DOCUMENTS STARTS =============================================================
\begin{document}
\begin{frame}
    \titlepage
\end{frame}
\begin{frame}{Overview}
    This talk will be based on the content of our draft paper and selected content of the Catalyst Meta Acceleration Framework. 
    Our preprint: 
    \begin{enumerate}
        \item H. Li and X. Wang, \textit{A Parameter Free Accelerated Proximal Gradient Method Without Restarting}, preprint, (2025).
    \end{enumerate}
    Catalyst Meta Acceleration:  
    \begin{enumerate}
        \item H. Lin, J. Mairal and Z. Harchaoui, \textit{A universal catalyst for first-order optimization}, in NISP, vol. 28, (2015). 
        \item \underline{\hspace{4em}}, \textit{Catalyst acceleration for first-order convex optimization: from theory to practice}, JMLR, 18 (2018), pp. 1–54.
    \end{enumerate}
\end{frame}

@@

        Part I: Introduction to Relaxed Weak Accelerated Proximal Gradient (R-WAPG). 

@@

                        y_k &= 

@@

                        v_{k + 1} &= 

@@

                        x_{k + 1} &= T_L y_k. 

@@

                \item Previous claimed generic convergence results will apply to all equivalent forms of R-WAPG will now come. 

@@

                        & y_{k} = 

@@

                        & x_{k + 1} = 

@@

                        & v_{k + 1} = 

@@

                        & y_k = 

@@

        \begin{frame}{R-WAPG momentum form}
            \begin{definition}[R-WAPG momentum form]\label{def:r-wapg-momentum-form}
                Given any $y_1 = x_1 \in \RR^n$, and sequences $(\rho_k)_{k \ge 0}, (\alpha_k)_{k\ge 0}$ Definition \ref{def:rwapg-seq}. 
                The algorithm generates iterates $x_{k + 1}, y_{k + 1}$ For $k = 1, 2, \ldots $ by the procedures: 
                {\small
                \begin{tcolorbox}
                    For $k=1, 2,\ldots $
                    \begin{align*}
                        & x_{k + 1} = y_k - L^{-1}\mathcal G_Ly_k, 
                        \\
                        & 
                        y_{k + 1} = 
                        x_{k + 1} + 
                        \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}}(x_{k + 1} - x_k). 
                    \end{align*}    
                \end{tcolorbox}
                }
                In the special case where $\mu = 0$, the momentum term can be represented without parameter $\rho_k$: 
                $$
                    (\forall k \ge 1)\quad \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}} 
                    = \alpha_{k + 1}(\alpha_k^{-1} - 1). 
                $$
            \end{definition}
        \end{frame}

@@

@@

                \item The synthetic solution is $x^+ = [1\; -1\; 1 \; \cdots ]^T \in \RR^N$ and $b = Ax^+ \in \RR^M$.

@@

                    A single LASSO experiment results, with $M = 64, 256$. 

@@

            We consider this a minor augmentation for the future. 

@@

                \item In Nesterov's 2013 paper \cite{nesterov_gradient_2013}, he considers accelerated  minimization problem $\phi = f + g$ with $f$ $L_f$ smooth and $g$ being $\mu_g \ge 0$ strongly convex. 
                \item Algorithm 5, Chambolle, Pock \cite{chambolle_introduction_2016} captures several variants of FISTA, and it assumes that $F = f + g$ where $f, g$ has strong convexity constant $\mu_f \ge 0, \mu_g \ge 0$ respective so $F$ is $\mu := \mu_f + \mu_g \ge 0$ strongly convex. 

@@

@@

                Suppose $\mathbb M$ generates iterates $(z_{k, t})_{t \ge 0}$ for the inner loop iteration such that there exists $C_{\mathbb M} > 0$, and it has: 

@@

                    \mathcal M_F^{\kappa^{-1}}(z_{k, t}, y) - \mathcal M^*_{F, \kappa^{-1}}(y) 

@@

                        \mathcal M_{F}^{\kappa^{-1}}(z_{k,0})

@@

                        \mathcal M^*_{F, \kappa^{-1}}(y)

@@

                Then the $\mathbb A$ generates $(x_{k})_{k \ge 0}$ such that 

@@

                        \left(1 + \frac{2}{\eta}\right)^2(F(x_k) - F^*)

@@

            We count $m$, the number of iteration experienced by $\mathbb M$ up to the $k$ iteration of $\mathbb A$. 

@@

                        \tau_{\mathbb M}\sqrt{\mu}/(\mu + \kappa)

@@

                    m &\le \sum_{i - 1}^{k} k T_{\mathbb M} \log(i + 2) \le k T_{\mathbb M} \log(k + 2) 

@@

                Let $(\delta_k)_{k \ge 0}$ be an absolute error sequence chosen for the evaluation for inexact proximal point method. 

@@

                    \item If $F$ is smooth then choose $z_k = y_k$. 

@@

                \item $N_{\mathbb N}\le \widetilde{\mathcal O}\left((\tau_{\mathbb M}\sqrt{q})^{-1}\log(\epsilon^{-1})\right)$ if $\mu > 0$. 

@@

        \end{frame}
        \begin{frame}{Our question on Catalyst Acceleration}
            \begin{block}{A key detail about Catalyst}
                The smoothness assumption is baked into Assumption \ref{ass:linear-convergence-inner-loop}, where it states that the $\mathbb M$ needs to have linear convergence rate. 
            \end{block}
            All VRMs method in the literature requires Lipschitz smoothness and strong convexity to achieve linear convergence rate. 
            No linear convergent first order algorithm exists for all strong convex function by Theorem 2.1.5 in Nesterov's book \cite{nesterov_lectures_2018}. 
            Hence, here is our question: 
            \begin{alertblock}{Our question}
                Solving inexact proximal point doesn't necessarily need smoothness assumption on $F$ so what if we Assumption \ref{ass:linear-convergence-inner-loop} is false? 
            \end{alertblock}
            Next slide we explain why the answer is not trivial and why it's interesting. 
        \end{frame}
        \begin{frame}{The answer to our question is nontrivial}
            Without smoothness assumption on $F$, the following considerations apply. 
            \begin{enumerate}
                \item $\mathbb A$ optimizes inexact Moreau Envelope $\mathcal M^*_{\kappa^{-1}, F}(y)$ which is smooth and strongly convex if $F$ is convex hence $\mathcal O(1/\sqrt{k})$ from Theorem 3.2.1 \cite{nesterov_lectures_2018} doesn't apply for the total complexity.  
                \item Theorem 2.1.7 \cite{nesterov_lectures_2018} doesn't apply because the lower complexity bound $\mathbb A$ is $\mathcal O(1/k^2)$ assume exact gradient evaluations. 
                \item Theorem 2.1.17 \cite{nesterov_lectures_2018} (Linear rate lower bound for strong convexity and smoothness) doesn't apply on $\mathbb M$. 
                Instead Theorem 3.2.5 applies and the lower bound is $\mathcal O(1/k)$. 
            \end{enumerate}
            Subgradient descent with Polyak step size reaches the lower bound of worst case of Strongly convex, strongly convex and smooth.
            See Hazan et al. \cite{hazan_revisiting_2022}.

>>>>

% DOCUMENTS STARTS =============================================================
\begin{document}
\begin{frame}
    \titlepage
\end{frame}
\begin{frame}{Overview}
    This talk will be based on the content of our draft paper and selected content of the Catalyst Meta Acceleration Framework. 
    Our preprint: 
    \begin{enumerate}
        \item H. Li and X. Wang, \textit{Relaxed Weak Accelerated Proximal Gradient Method: a Unified Framework for Nesterov's Accelerations}, preprint, (2025).
    \end{enumerate}
    Catalyst Meta Acceleration:  
    \begin{enumerate}
        \item H. Lin, J. Mairal and Z. Harchaoui, \textit{A universal catalyst for first-order optimization}, in NISP, vol. 28, (2015). 
        \item \underline{\hspace{4em}}, \textit{Catalyst acceleration for first-order convex optimization: from theory to practice}, JMLR, 18 (2018), pp. 1–54.
    \end{enumerate}
\end{frame}

@@

        Part I: Introduction to Relaxed Weak Accelerated Proximal Gradient (R-WAPG) and Catalyst Acceleration. 

@@

                        y_k &:= 

@@

                        v_{k + 1} &:= 

@@

                        x_{k + 1} &:= T_L y_k. 

@@

                \item Previous claimed generic convergence results will apply to all equivalent forms of R-WAPG. 

@@

                        & y_{k} := 

@@

                        & x_{k + 1} := 

@@

                        & v_{k + 1} := 

@@

                        & y_k := 

@@

        \begin{frame}{R-WAPG momentum form}
            \begin{definition}[R-WAPG momentum form]\label{def:r-wapg-momentum-form}
                Given any $y_1 = x_1 \in \RR^n$, and sequences $(\rho_k)_{k \ge 0}, (\alpha_k)_{k\ge 0}$ Definition \ref{def:rwapg-seq}. 
                The algorithm generates iterates $x_{k + 1}, y_{k + 1}$ For $k = 1, 2, \ldots $ by the procedures: 
                {\small
                \begin{tcolorbox}
                    For $k=1, 2,\ldots $
                    \begin{align*}
                        & x_{k + 1} := y_k - L^{-1}\mathcal G_Ly_k, 
                        \\
                        & 
                        y_{k + 1} := 
                        x_{k + 1} + 
                        \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}}(x_{k + 1} - x_k). 
                    \end{align*}    
                \end{tcolorbox}
                }
                In the special case where $\mu = 0$, the momentum term can be represented without parameter $\rho_k$: 
                $$
                    (\forall k \ge 1)\quad \frac{\rho_k\alpha_k(1 - \alpha_k)}{\rho_k\alpha_k^2 + \alpha_{k + 1}} 
                    = \alpha_{k + 1}(\alpha_k^{-1} - 1). 
                $$
            \end{definition}
        \end{frame}

@@

                        \mu, 

@@

                \item The synthetic solution is $x^+ = [1\;, -1\;, 1 \;, \cdots ]^T \in \RR^N$ and $b = Ax^+ \in \RR^M$.

@@

                    A single LASSO experiment results, with $M = 64, N = 256$. 

@@

            We consider this a minor augmentation in the future. 

@@

                \item In Nesterov's 2013 paper \cite{nesterov_gradient_2013}, he considered accelerated minimization problem $\phi = f + g$ where $f$ is $L_f$ smooth and $g$ is $\mu_g \ge 0$ strongly convex. 
                \item Algorithm 5, Chambolle and Pock \cite{chambolle_introduction_2016} captured some variants of FISTA, and they assumed that $F = f + g$ where $f, g$ has strong convexity constant $\mu_f \ge 0, \mu_g \ge 0$ respective so $F$ is $\mu := \mu_f + \mu_g \ge 0$ strongly convex. 

@@

            Notations for inner loop, outer loop of Catalyst Acceleration. 

@@

                Suppose $\mathbb M$ generates iterates $(z_{k, t})_{t \ge 0}$ for the inner loop iteration such that there exists $C_{\mathbb M} > 0, \tau_{\mathbb M} \in (0, 1)$, and it has: 

@@

                    \mathcal M_F^{1/\kappa}(z_{k, t}, y) - \mathcal M^*_{F, 1/\kappa}(y) 

@@

                        \mathcal M_{F}^{1/\kappa}(z_{k,0})

@@

                        \mathcal M^*_{F, 1/\kappa}(y)

@@

                Then $\mathbb A$ generates $(x_{k})_{k \ge 0}$ such that 

@@

                        \left(1 + \frac{2}{\eta}\right)^2(F(x_0) - F^*)

@@

            We count $m$, the number of total iterations experienced by $\mathbb M$ up to the $k$ iteration of $\mathbb A$. 

@@

                            1 - \tau_{\mathbb M}\sqrt{\mu/(\mu + \kappa)}
                        \right)^m

@@

                    m 
                    \sum_{i = 1}^{k} T_{\mathbb M} \log(i + 2) 
                        \le k T_{\mathbb M} \log(k + 2) 

@@

                Let $(\delta_k)_{k \ge 0}$ be an relative error sequence chosen for the evaluation for inexact proximal point method. 

@@

                    \item If $F$ is smooth then choose $z_{k, 0} = y_k$. 

@@

                \item $N_{\mathbb M}\le \widetilde{\mathcal O}\left((\tau_{\mathbb M}\sqrt{q})^{-1}\log(\epsilon^{-1})\right)$ if $\mu > 0$. 

@@

        \end{frame}
        \begin{frame}{Our question on Catalyst Acceleration}
            \begin{block}{A key detail about Catalyst}
                The smoothness assumption is used in Assumption \ref{ass:linear-convergence-inner-loop}, where it states that the $\mathbb M$ needs to have linear convergence rate. 
            \end{block}
            All VRMs method in the literature requires Lipschitz smoothness and strong convexity to achieve linear convergence rate. 
            No linear convergent first order algorithm exists for all strongly convex function by Theorem 2.1.5 in Nesterov's book \cite{nesterov_lectures_2018}. 
            Hence, here is our question: 
            \begin{alertblock}{Our question}
                Solving inexact proximal point doesn't necessarily need smoothness assumption on $F$ so what if we Assumption \ref{ass:linear-convergence-inner-loop} is false? 
            \end{alertblock}
            Next slide we explain why the answer is not trivial and why it's interesting. 
        \end{frame}
        \begin{frame}{The answer to our question is nontrivial}
            Without smoothness assumption on $F$, the following considerations apply. 
            \begin{enumerate}
                \item $\mathbb A$ optimizes inexact Moreau Envelope $\mathcal M^*_{\kappa^{-1}, F}(y)$ which is smooth and strongly convex if $F$ is convex hence $\mathcal O(1/\sqrt{k})$ from Theorem 3.2.1 \cite{nesterov_lectures_2018} doesn't apply for the total complexity.  
                \item Theorem 2.1.7 \cite{nesterov_lectures_2018} doesn't apply because the lower complexity bound $\mathbb A$ is $\mathcal O(1/k^2)$ assume exact gradient evaluations. 
                \item Theorem 2.1.17 \cite{nesterov_lectures_2018} (Linear rate lower bound for strong convexity and smoothness) doesn't apply on $\mathbb M$. 
                Instead, Theorem 3.2.5 applies and the lower bound is $\mathcal O(1/k)$. 
            \end{enumerate}
            Subgradient descent with Polyak step size reaches the lower bound of worst case of Strongly convex, strongly convex and smooth.
            See Hazan et al. \cite{hazan_revisiting_2022}.
